---
title: 'Project 1: Shot Chart Data (No Free Throws Accounted)'
author: "Austin Shih"
date: '2022-09-27'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

- Shot Chart of Kevin Durant in the 2018-2019 Season 
```{r}
require(tidyverse)
require(dplyr)

nba <- read_csv('../data/nba_shotchartdetail_2018-19.csv')
glimpse(nba)
nba %>% 
  filter(PLAYER_NAME == 'Kevin Durant') %>%
  select(LOC_X,LOC_Y,SHOT_MADE_FLAG) %>%
  ggplot(aes(x = LOC_X, y = LOC_Y, color = factor(SHOT_MADE_FLAG))) + 
  geom_point(alpha = 0.5) + 
  labs(title = "Kevin Durant Shot Chart", subtitle = "2018-19 Season") + 
  geom_curve(x = -200, y = 0, xend = 200, yend = 0, curvature = -1.2, color = 'black') +
  scale_color_manual(name = 'Make or Miss', 
                     values = c('red', 'green'), 
                     labels = c('Miss', 'Make'))
```

- Who had the highest shooting percentage among players who shot more than 500 shots in the season? 
```{r}
nba %>%
  group_by(PLAYER_NAME) %>%
  count(EVENT_TYPE) %>%
  mutate(shooting_pct = n/sum(n)) %>%
  filter(sum(n) > 500) %>%
  filter(EVENT_TYPE == 'Made Shot') %>%
  arrange(-shooting_pct) 
```
> Rudy Gobert had the highest shooting percentage at 66.9% 

- League Average 
```{r}
league <- nba %>%
  count(EVENT_TYPE) %>%
  mutate(league_avg = n/sum(n)) %>%
  filter(EVENT_TYPE == 'Made Shot') %>%
  select(league_avg)

player <- nba %>%
  group_by(PLAYER_NAME) %>%
  count(EVENT_TYPE) %>%
  mutate(shooting_pct = n/sum(n)) %>%
  filter(EVENT_TYPE == 'Made Shot') %>%
  arrange(-shooting_pct) 

shooting <- NULL
shooting <- cbind(player, league) 

shooting %>%
  ggplot(aes(x = shooting_pct, y = n)) + 
  geom_point(alpha = 0.5) + 
  geom_vline(xintercept = 0.4605072, color = 'red') + 
  labs(title = 'Player Shooting Average vs NBA Shooting Average', 
       x = 'Player - NBA Shooting Average', 
       y = 'Number of Made Shots')
```
> The more shots a player makes, the closer a player's shooting average is to the NBA shooting average. 

- Highest Point Scorers Per Month
```{r}
points <- nba %>%
  mutate(GAME_DATE = as.Date(factor(GAME_DATE), format = '%Y%m%d')) %>%
  group_by(PLAYER_NAME) %>%
  mutate(points_scored = 0) %>%
  filter(EVENT_TYPE == 'Made Shot') %>%
  mutate(points_scored = ifelse(SHOT_TYPE == '3PT Field Goal', points_scored + 3,
                                ifelse(SHOT_TYPE == '2PT Field Goal', points_scored + 2, points_scored + 0))) %>%
  ungroup()

points %>%
  group_by(PLAYER_NAME) %>%
  mutate(season_points = sum(points_scored)) %>%
  select(PLAYER_NAME, season_points) %>%
  distinct() %>%
  arrange(-season_points)

points <- points %>%
  group_by(PLAYER_NAME,GAME_ID) %>%
  mutate(game_points = sum(points_scored)) %>%
  ungroup()

points %>%
  select(GAME_ID, PLAYER_NAME, GAME_DATE, game_points) %>%
  distinct() %>% 
  group_by(PLAYER_NAME,month = lubridate::floor_date(GAME_DATE, 'month')) %>%
  mutate(month_points = sum(game_points)) %>%
  select(PLAYER_NAME, month, month_points) %>%
  distinct() %>%
  arrange(-month_points)
```
> James Harden had the most amount of points scored in a month with 436 points in the month of January 2019. 

- Average PPG for James Harden by month 
```{r}
points %>%
  filter(PLAYER_NAME == 'James Harden') %>%
  select(GAME_ID, PLAYER_NAME, GAME_DATE, game_points) %>%
  distinct() %>% 
  group_by(month = lubridate::floor_date(GAME_DATE, 'month')) %>%
  mutate(month_points = sum(game_points)) %>%
  select(PLAYER_NAME, month, month_points) %>%
  distinct() %>%
  ggplot(aes(x = month, y = month_points)) + 
  geom_bar(stat = 'identity') + 
  labs(title = 'James Harden - Total Points', 
       subtitle = 'Not Counting Free Throws',
       x = 'Month', 
       y = 'Points in a Month')

points %>%
  filter(PLAYER_NAME == 'James Harden') %>%
  select(GAME_ID, PLAYER_NAME, GAME_DATE, game_points) %>%
  distinct() %>%
  group_by(month = lubridate::floor_date(GAME_DATE, 'month')) %>%
  mutate(avg_month_points = mean(game_points)) %>%
  select(PLAYER_NAME, month, avg_month_points) %>%
  distinct() %>%
  ggplot(aes(x = month, y = avg_month_points)) + 
  geom_bar(stat = 'identity') + 
  labs(title = 'James Harden - PPG', 
       subtitle = 'Not Counting Free Throws',
       x = 'Month', 
       y = 'Average PPG per month')
```

- James Harden Lowest Shooting % by Action
```{r}
nba %>%
  filter(PLAYER_NAME == 'James Harden') %>%
  select(PLAYER_NAME, EVENT_TYPE, ACTION_TYPE) %>%
  mutate(bucket = 0) %>%
  mutate(bucket = ifelse(EVENT_TYPE == 'Made Shot', bucket + 1, 0)) %>%
  group_by(ACTION_TYPE) %>%
  mutate(avg_per_shot = mean(bucket)) %>%
  filter(EVENT_TYPE == 'Made Shot') %>%
  distinct() %>%
  arrange(avg_per_shot) %>%
  head() %>%
  ggplot(aes(x = avg_per_shot, y = ACTION_TYPE)) + 
  geom_bar(stat = 'identity') + 
  labs(title = 'James Harden: Lowest Shooting % per Action', 
       x = 'Shooting %', 
       y = '') +
  scale_x_continuous(labels = scales::percent)
  
```
